\bf Tearing Mode Dynamics in Tokamak Plasmas

### 2021-10-30 Documentation/TearingModeTheory.tex
Added to tearing mode writeup

\begin{document}
\title{\bf Tearing Modes in Tokamak Plasmas}
\author{R.~Fitzpatrick\,\footnote{rfitzp@utexas.edu}\\
Institute for Fusion Studies,\\  Department of Physics,\\  University of Texas at Austin,\\  Austin TX 78712, USA}
\maketitle

\section{Fluid Theory}
\subsection{Fundamental Quantities}
Consider an idealized tokamak plasma consisting of an equal number of electrons, with
mass $m_e$ and charge $-e$ (here, $e$ denotes the  magnitude of the electron
charge), and ions, with  mass $m_i$ and charge $+e$. We shall employ 
the symbol
\begin{equation}
T_s = \frac{1}{3} \,m_s\, \langle v_s^{2}\rangle
\end{equation}
to denote a {\em kinetic temperature}\/   measured in units of energy.
Here, $v$ is a particle speed, and the angular brackets denote an
ensemble average (Reif 1965). The kinetic temperature of species $s$ is a measure of 
the mean kinetic energy of particles of
that species. (Here, $s$ represents either $e$ for electrons, or $i$ for ions.) In tokamak plasma physics, kinetic temperature
is invariably measured in {\em electron-volts}\/ \index{electron-volt} (1 joule is equivalent to
$6.24\times 10^{18}$\,eV).

@@

In a tokamak plasma,  the ambient magnetic field, 
${\bf B}$,  is strong enough to significantly alter charged particle
trajectories. In particular,
magnetized plasmas are highly  anisotropic, responding differently to
forces that are parallel and perpendicular to the direction
of ${\bf B}$. As is well known, charged particles respond to the Lorentz force
by freely streaming in the direction of ${\bf B}$, while executing
circular  {\em gyro-orbits} in the plane perpendicular to ${\bf B}$ (Fitzpatrick 2008). 
The typical   {\em gyroradius}\/ of a charged particle 
gyrating in a
magnetic field is given by
\begin{equation}
\rho_s = \frac{v_{t\,s}}{{\mit\Omega}_s},
\end{equation}
where
\begin{equation}
{\mit\Omega}_s = \frac{e\,B}{m_s}
\end{equation}
is the  {\em gyrofrequency}\/ associated with the
gyration.

@@

\subsection{Braginskii Equations}
The electron and ion fluid equations in a collisional plasma
take the form:
\begin{align}\label{e3.79a}
\frac{dn}{dt} + n\,\nabla\cdot{\bf V}_e &= 0,\\[0.5ex]
m_e \,n\,\frac{d {\bf V}_e}{dt} + \nabla p_e+ \nabla\cdot \bpi_e + e\, n\,
({\bf E} + {\bf V}_e\times {\bf B})& = {\bf F},\\[0.5ex]
\frac{3}{2}\frac{d p_e}{dt} + \frac{5}{2}\,p_e\,\nabla\cdot{\bf V}_e
+ \bpi_e:\nabla{\bf V}_e+ \nabla\cdot{\bf q}_e &= w_e,\label{e3.79c}
\end{align}
and
\begin{align}\label{e3.80a}
\frac{dn}{dt} + n\,\nabla\cdot{\bf V}_i &= 0,\\[0.5ex]
m_i \,n\,\frac{d {\bf V}_i}{dt} + \nabla p_i + \nabla\cdot \bpi_i - e\, n\,
({\bf E} + {\bf V}_i\times {\bf B})& =- {\bf F},\\[0.5ex]
\frac{3}{2}\frac{d p_i}{dt} + \frac{5}{2}\,p_i\,\nabla\cdot{\bf V}_i
+ \bpi_i:\nabla{\bf V}_i+ \nabla\cdot{\bf q}_i &= w_i,\label{e3.80c}
\end{align}
respectively. 
Equations (\ref{e3.79a})--(\ref{e3.79c}) and (\ref{e3.80a})--(\ref{e3.80c})  are called the {\em Braginskii equations}, because they were first obtained
in a celebrated article by S.I.~Braginskii (Braginskii 1965). Here, ${\bf V}_s$ is the species-$s$ flow velocity, $p_s=n\,T_s$ is the 
species-$s$ scalar
pressure, $\bpi_s$ is the species-$s$ viscosity tensor, ${\bf F}$ is the collisional friction force density, ${\bf q}_s$ is the species-$s$ heat flux density, $w_s$ is the species-$s$ collisional heating rate density, and ${\bf E}$ is the ambient electric field strength. Moreover, $d/dt\equiv \partial/\partial t + {\bf V}_e\cdot\nabla$ in the electron fluid equations, (\ref{e3.79a})--(\ref{e3.79c}), whereas  $d/dt\equiv \partial/\partial t + {\bf V}_i\cdot\nabla$ in the ion fluid equations,  (\ref{e3.80a})--(\ref{e3.80c}).

@@

In order to describe the viscosity tensor in a magnetized plasma, it is
helpful to define the {\em rate-of-strain tensor} \index{rate-of-strain tensor} 
\begin{equation}
W_{\alpha\beta} = \frac{\partial V_\alpha}{\partial r_\beta}
+ \frac{\partial V_\beta}{\partial r_\alpha} - \frac{2}{3} \,\nabla\cdot{\bf V}\,
\delta_{\alpha\beta}.
\end{equation}
Obviously, there is a separate rate-of-strain tensor for the electron and ion
fluids. It is easily demonstrated that this tensor is zero if the plasma
translates, or rotates as a rigid body, or if it undergoes isotropic
compression. Thus, the rate-of-strain tensor measures the  deformation  of
plasma volume elements.

@@

\subsection{Lowest-Order Fluid Equations}
If we restore dimensions to the Braginskii equations then we can write them in the form
\begin{align}
\frac{\partial n}{\partial t} + \nabla\cdot(n\,{\bf V}_e) &= 0,\label{e3.151a}\\[0.5ex]
m_e\, n\,\frac{\partial  {\bf V}_e}{\partial t} +
m_e\, n\,({\bf V}_e\cdot\nabla){\bf V}_e+
 [\delta_e^{-2}]\,\nabla p_e+
[\zeta_e^{-1}]\, \nabla\cdot \bpi_e &\label{e3.151b}\\[0.5ex]
 +[\delta_e^{-2}]\,e\, n\,
({\bf E} + {\bf V}_e\times {\bf B})& =[\delta_e^{-2}\,\zeta_e]\, {\bf F}_U + [\delta_e^{-2}]\,
{\bf F}_T,\nonumber\\[0.5ex]
\frac{3}{2}\frac{\partial  p_e}{\partial t} + \frac{3}{2}
 \,({\bf V}_e\cdot\nabla)\, p_e 
+ \frac{5}{2}\,p_e\,\nabla\cdot{\bf V}_e+[\delta_e^{\,2}\,\zeta_e^{-1}]\, \bpi_e:\nabla{\bf V}_e&\label{e3.151c}\\[0.5ex]
+ [\zeta_e^{-1}]\,\nabla\cdot{\bf q}_{Te} +\nabla\cdot{\bf q}_{Ue}  &=-[\delta_e^{-2}\,\zeta_e\,\mu^2]\,
 w_i+ [\zeta_e]\,w_U + w_T,\nonumber
\end{align}
and
\begin{align}\label{e3.152a}
\frac{\partial n}{\partial t} +\nabla\cdot(n\,{\bf V}_i) &= 0,\\[0.5ex]
m_i \,n\,\frac{\partial  {\bf V}_i}{\partial t} +
m_i \,n\,({\bf V}_i\cdot\nabla) {\bf V}_i +
 [\delta_i^{-2}]\,\nabla p_i + 
[\zeta_i^{-1}]\,\nabla\cdot \bpi_i &\label{e3.152b}\\[0.5ex]
- [\delta_i^{-2}]\,e \,n\,
({\bf E} + {\bf V}_i\times {\bf B})& =-[\delta_i^{-2}\,\zeta_i\,\mu]\, {\bf F}_U - [\delta_i^{-2}]\,{\bf F}_T,
\nonumber\\[0.5ex]
\frac{3}{2}\frac{\partial  p_i}{\partial t} +   \frac{3}{2}
 \,({\bf V}_i\cdot\nabla) \,p_i
+ \frac{5}{2}\,p_i\,\nabla\cdot{\bf V}_i+[\delta_i^{\,2}\,\zeta_i^{-1}]\, \bpi_i:\nabla{\bf V}_i\nonumber\\[0.5ex]
+ [\zeta_i^{-1}]\,\nabla\cdot{\bf q}_i &= [\delta_i^{-2}\,\zeta_i\,\mu]\,w_i.\label{e3.152c}
\end{align}
The terms in square brackets are just to remind us that the terms they precede are smaller or larger than the
other terms in the equations (by the corresponding terms within the brackets).

If we assume that $\delta_e\sim \zeta_e\sim \mu$ then the dominant terms in the electron energy conservation equation
(\ref{e3.151c}) yield
\begin{equation}
\nabla\cdot{\bf q}_{T_e}\simeq \nabla\cdot(-\kappa_\parallel^e\,\nabla_\parallel T_e)\simeq 0,
\end{equation}
which implies that 
\begin{equation}
{\bf B}\cdot\nabla T_e=0. \label{e75}
\end{equation}
In other words, the parallel electron heat conductivity in a tokamak plasma is usually
sufficiently large that it forces the electron temperature to be constant on magnetic flux-surfaces. 
Likewise, if we assume that $\delta_i\sim \zeta_i\sim \mu$ then the dominant terms in the ion energy conservation equation
(\ref{e3.152c}) yield
\begin{equation}
\nabla\cdot{\bf q}_{T_i}\simeq \nabla\cdot(-\kappa_\parallel^i\,\nabla_\parallel T_i)\simeq 0,
\end{equation}
which implies that 
\begin{equation}
{\bf B}\cdot\nabla T_i=0.\label{e77}
\end{equation}
In other words, the parallel ion heat conductivity in a tokamak plasma is usually
sufficiently large that it forces the ion temperature to be constant on magnetic flux-surfaces.

 The dominant terms in the electron and ion momentum conservation equations,
(\ref{e3.151b}) and (\ref{e3.152b}), respectively, yield
\begin{align}\label{e78}
\nabla p_e + n\,e\,({\bf E}+{\bf V}_e\times {\bf B}) &\simeq 0,\\[0.5ex]
\nabla p_i - n\,e\,({\bf E}+{\bf V}_i\times {\bf B}) &\simeq 0.\label{e79}
\end{align}
The sum of the preceding two equations gives
\begin{equation}\label{e80}
 {\bf j}\times {\bf B} \simeq\nabla p,
\end{equation}
where $p=p_e+p_i$. In other words, a conventional tokamak plasma exists in a state in which electromagnetic
forces exactly balance the  total scalar pressure force. 
It follows from the previous equation that 
\begin{equation}
{\bf B}\cdot\nabla p\simeq 0.
\end{equation}
In other words, the pressure in a conventional tokamak plasma is constant on magnetic flux-surfaces. 
 However, given that $p=n\,(T_e+T_i)$, and making use of Eqs.~(\ref{e75}) and (\ref{e77}), we deduce that 
\begin{equation}
{\bf B}\cdot\nabla n\simeq 0.
\end{equation}
In other words, the particle density in a conventional tokamak plasma is also  constant on
magnetic flux-surfaces.

 Taking the difference of the electron and ion particle
conservation equations, (\ref{e3.151a}) and (\ref{e3.152a}), respectively, we obtain
\begin{equation}\label{e82}
\nabla\cdot {\bf j} = 0.
\end{equation}
 Equations~(\ref{e80}) and (\ref{e82}) can be combined to give
\begin{equation}\label{e84}
({\bf B}\cdot\nabla)\,{\bf j } - ({\bf j}\cdot\nabla)\,{\bf B} \simeq {\bf 0}.
\end{equation}

Finally, Equations~(\ref{e78}) and (\ref{e79}) imply that
\begin{align}
{\bf V}_{\perp\,e} &\simeq {\bf V}_E + {\bf V}_{\ast\,e},\\[0.5ex]
{\bf V}_{\perp\,i} &\simeq {\bf V}_E + {\bf V}_{\ast\,i},
\end{align}
where
\begin{equation}
{\bf V}_e = \frac{{\bf E}\times {\bf B}}{B^2}
\end{equation}
is the {\em E-cross-B drift velocity}, and
\begin{align}
{\bf V}_{\ast\,e}&= \frac{\nabla p_e\times {\bf B}}{e\,n\,B^2},\\[0.5ex]
{\bf V}_{\ast\,i}&=- \frac{\nabla p_i\times {\bf B}}{e\,n\,B^2},
\end{align}
are termed the {\em electron diamagnetic velocity}\/ and the {\em ion diamagnetic velocity}, respectively.

\section{Cylindrical Tearing Mode Theory}
\subsection{Tokamak Equilibrium}
Consider a large aspect-ratio, toroidal,  tokamak plasma equilibrium whose magnetic flux-surfaces map out (almost) concentric circles in the poloidal plane. 
Let us employ a conventional set of right-handed cylindrical coordinates, $r$, $\theta$, $z$. The
equilibrium magnetic flux-surfaces lie on surfaces of constant $r$. The system is
assumed to be periodic in the $z$ (`toroidal') direction, with periodicity length $2\pi\,R_0$, 
where $R_0$ is the simulated {\em major radius}\/ of the plasma. The {\em safety-factor}\/ profile takes the form
\begin{equation}\label{e90}
q(r)= \frac{r\,B_z}{R_0\,B_\theta(r)},
\end{equation}
where $B_z$ is the constant `toroidal' magnetic field-strength, and $B_\theta(r)$ is
the poloidal magnetic field-strength. It is assumed that $q\sim{\cal O}(1)$. 
The equilibrium `toroidal' current density
satisfies
\begin{equation}
\mu_0\,j_z(r) = \frac{(r\,B_\theta)'}{r},
\end{equation}
where $'$ denotes $d/dr$. Finally, the standard large aspect-ratio
tokamak orderings,  
\begin{equation}\label{e92}
\frac{r}{R_0}\ll 1,~~~ \frac{B_\theta}{B_z}\ll 1,
\end{equation}
are adopted (Fitzpatrick 1993).

\subsection{Perturbed Magnetic Field}
Consider a tearing mode perturbation that has $m$ periods in the poloidal direction,
and $n$ periods in the `toroidal' direction, where $m/n\sim {\cal O}(1)$.  We can write the perturbed magnetic field in the form
\begin{equation}\label{e93}
\delta{\bf B} \simeq \nabla\times (\psi\,{\bf e}_z),
\end{equation}
where
\begin{equation}\label{e93a}
\psi(r,\theta,\phi,t)= \psi(r,t)\,\exp[{\rm i}\,(m\,\theta-n\,\phi)].
\end{equation}
Here, $\phi=z/R_0$ is a simulated toroidal angle. The perturbed current density becomes
\begin{equation}\label{e94}
\mu_0\,\delta {\bf j} \simeq - \nabla^2\psi\,{\bf e}_z.
\end{equation}
It is assumed that $|\delta{\bf B}|\ll |{\bf B}|$.

\subsection{Cylindrical Tearing Mode Equation}
To lowest order, the tearing perturbation is governed by the linearized version of Equation~(\ref{e84}), which yields
\begin{equation}\label{e96}
(\delta{\bf B}\cdot \nabla)\,{\bf j} + ({\bf B}\cdot\nabla)\,\delta{\bf j} - (\delta{\bf j}\cdot\nabla)\,{\bf B} - 
({\bf j}\cdot\nabla)\,\delta {\bf B}\simeq {\bf 0}.
\end{equation}
Making use of Equations~(\ref{e90})--(\ref{e94}), the $z$-component of the previous equation gives
the {\em cylindrical tearing mode equation}\/ (Fitzpatrick 1993),
\begin{equation}
\frac{\partial^2\psi}{dr^2} + \frac{1}{r}\,\frac{\partial\psi}{\partial r}-\frac{m^2}{r^2}\,\psi - \frac{J_z'\,\psi}{r\,(1/q-n/m)}\simeq 0,
\end{equation}
where 
\begin{equation}
J_z(r)= \frac{R_0\,\mu_0\,j_z(r)}{B_z}.
\end{equation}

\subsection{Solution in Presence of Perfectly-Conducting Wall}\label{perfect}
Suppose that the plasma occupies the region $0\leq r\leq a$, where $a$ is the plasma {\em minor radius}. It follows that
$J_z(r)=0$ for $r>a$. Suppose that the plasma is surrounded by a concentric perfectly-conducting wall of radius $r_w>a$. An appropriate
physical solution of the cylindrical tearing mode equation takes the form
\begin{equation}\label{e99}
\psi(r,t)= {\mit\Psi}_s(t)\,\hat{\psi}_s(r),
\end{equation}
where the real function $\hat{\psi}_s(r)$ is a solution of 
\begin{equation}\label{e100}
\frac{d^2\hat{\psi}_s}{dr^2} + \frac{1}{r}\,\frac{d\hat{\psi}_s}{dr}-\frac{m^2}{r^2}\,\hat{\psi}_s - \frac{J_z'\,\hat{\psi}_s}{r\,(1/q-n/m)}\simeq 0
\end{equation}
that satisfies
\begin{align}\label{e101}
\hat{\psi}_s(0) &= 0,\\[0.5ex]
\hat{\psi}_s(r_s) &= 1,\label{e102}\\[0.5ex]
\hat{\psi}_s(r\geq r_w) &= 0.\label{e103}
\end{align}

>>>>

\begin{document}
\title{\bf Tearing Mode Dynamics in Tokamak Plasmas}
\author{R.~Fitzpatrick\,\footnote{rfitzp@utexas.edu}\\
Institute for Fusion Studies,\\  Department of Physics,\\  University of Texas at Austin,\\  Austin TX 78712, USA}
\maketitle

\section{Fluid Theory}
\subsection{Introduction}
The aim of this section is to introduce the fundamental plasma fluid theory that underpins the analysis of tearing mode dynamics in tokamak
plasmas.

\subsection{Fundamental Quantities}
Consider an idealized tokamak plasma consisting of an equal number of electrons, with
mass $m_e$ and charge $-e$ (here, $e$ denotes the  magnitude of the electron
charge), and ions, with  mass $m_i$ and charge $+e$. We shall employ 
the symbol
\begin{equation}
T_s = \frac{1}{3} \,m_s\, \langle v_s^{2}\rangle
\end{equation}
to denote a {\em kinetic temperature}\/   measured in units of energy.
Here, $v$ is a particle speed, and the angular brackets denote an
ensemble average (Reif 1965). The kinetic temperature of species $s$ is a measure of 
the mean kinetic energy of particles of
that species. (Here, $s$ represents either $e$ for electrons, or $i$ for ions.) In tokamak plasma physics, kinetic temperature
is invariably measured in {\em electron-volts}\/ \index{electron-volt} (1 joule is equivalent to
$6.24\times 10^{18}$\,eV).

@@

In a tokamak plasma,  the ambient magnetic field, 
${\bf B}$,  is strong enough to significantly alter charged particle
trajectories. Consequently,
tokamak plasmas are highly  anisotropic, responding differently to
forces that are parallel and perpendicular to the direction
of ${\bf B}$. As is well known, charged particles respond to the Lorentz force
by freely streaming in the direction of ${\bf B}$, while executing
circular  {\em gyro-orbits} in the plane perpendicular to ${\bf B}$ (Fitzpatrick 2008). 
The typical   {\em gyroradius}\/ of a charged particle 
gyrating in a
magnetic field is given by
\begin{equation}
\rho_s = \frac{v_{t\,s}}{{\mit\Omega}_s},
\end{equation}
where
\begin{equation}
{\mit\Omega}_s = \frac{e\,B}{m_s}
\end{equation}
is the  {\em gyrofrequency}\/ associated with the
gyration.

@@

\subsection{Braginskii Equations}\label{sbrag}
The electron and ion fluid equations in a collisional plasma
take the form:
\begin{align}\label{e3.79a}
\frac{dn}{dt} + n\,\nabla\cdot{\bf V}_e &= 0,\\[0.5ex]
m_e \,n\,\frac{d {\bf V}_e}{dt} + \nabla p_e+ \nabla\cdot \bpi_e + e\, n\,
({\bf E} + {\bf V}_e\times {\bf B})& = {\bf F},\label{e3.79b}\\[0.5ex]
\frac{3}{2}\frac{d p_e}{dt} + \frac{5}{2}\,p_e\,\nabla\cdot{\bf V}_e
+ \bpi_e:\nabla{\bf V}_e+ \nabla\cdot{\bf q}_e &= w_e,\label{e3.79c}
\end{align}
and
\begin{align}\label{e3.80a}
\frac{dn}{dt} + n\,\nabla\cdot{\bf V}_i &= 0,\\[0.5ex]
m_i \,n\,\frac{d {\bf V}_i}{dt} + \nabla p_i + \nabla\cdot \bpi_i - e\, n\,
({\bf E} + {\bf V}_i\times {\bf B})& =- {\bf F},\label{e3.80b}\\[0.5ex]
\frac{3}{2}\frac{d p_i}{dt} + \frac{5}{2}\,p_i\,\nabla\cdot{\bf V}_i
+ \bpi_i:\nabla{\bf V}_i+ \nabla\cdot{\bf q}_i &= w_i,\label{e3.80c}
\end{align}
respectively. 
Equations (\ref{e3.79a})--(\ref{e3.80c})  are called the {\em Braginskii equations}, because they were first obtained
in a celebrated article by S.I.~Braginskii (Braginskii 1965). Here, $n$ is the electron number density, ${\bf V}_s$ is the species-$s$ flow velocity, $p_s=n\,T_s$ is the 
species-$s$ scalar
pressure, $\bpi_s$ is the species-$s$ viscosity tensor, ${\bf F}$ is the collisional friction force density, ${\bf q}_s$ is the species-$s$ heat flux density, $w_s$ is the species-$s$ collisional heating rate density, and ${\bf E}$ is the ambient electric field strength. Moreover, $d/dt\equiv \partial/\partial t + {\bf V}_e\cdot\nabla$ in the electron fluid equations, (\ref{e3.79a})--(\ref{e3.79c}), whereas  $d/dt\equiv \partial/\partial t + {\bf V}_i\cdot\nabla$ in the ion fluid equations,  (\ref{e3.80a})--(\ref{e3.80c}).

@@

In order to describe the viscosity tensor in a magnetized plasma, it is
helpful to define the {\em rate-of-strain tensor} \index{rate-of-strain tensor} 
\begin{equation}
W_{\alpha\beta} = \frac{\partial V_\alpha}{\partial r_\beta}
+ \frac{\partial V_\beta}{\partial r_\alpha} - \frac{2}{3} \,\nabla\cdot{\bf V}\,
\delta_{\alpha\beta}.
\end{equation}
Obviously, there is a separate rate-of-strain tensor for the electron and ion
fluids. It is easily demonstrated that this tensor is zero if the fluid
translates, or rotates as a rigid body, or if it undergoes isotropic
compression. Thus, the rate-of-strain tensor measures the  deformation  of
fluid volume elements.

@@

\subsection{Lowest-Order Fluid Equations}\label{lowest}
If we restore dimensions to the Braginskii equations then we can write them in the form
\begin{align}
\frac{\partial n}{\partial t} + \nabla\cdot(n\,{\bf V}_e) &= 0,\label{e3.151a}\\[0.5ex]
m_e\, n\,\frac{\partial  {\bf V}_e}{\partial t} +
m_e\, n\,({\bf V}_e\cdot\nabla){\bf V}_e+
 [\delta_e^{-2}]\,\nabla p_e+
[\zeta_e^{-1}]\, \nabla\cdot \bpi_e &\nonumber\\[0.5ex]
 +[\delta_e^{-2}]\,e\, n\,
({\bf E} + {\bf V}_e\times {\bf B})& =[\delta_e^{-2}\,\zeta_e]\, {\bf F}_U + [\delta_e^{-2}]\,
{\bf F}_T,\label{e3.151b}\\[0.5ex]
\frac{3}{2}\frac{\partial  p_e}{\partial t} + \frac{3}{2}
 \,{\bf V}_e\cdot\nabla p_e 
+ \frac{5}{2}\,p_e\,\nabla\cdot{\bf V}_e+[\delta_e^{\,2}\,\zeta_e^{-1}]\, \bpi_e:\nabla{\bf V}_e&\label{e3.151c}\\[0.5ex]
+ [\zeta_e^{-1}]\,\nabla\cdot{\bf q}_{Te} +\nabla\cdot{\bf q}_{Ue}  &=-[\delta_e^{-2}\,\zeta_e\,\mu^2]\,
 w_i+ [\zeta_e]\,w_U + w_T,\nonumber
\end{align}
and
\begin{align}\label{e3.152a}
\frac{\partial n}{\partial t} +\nabla\cdot(n\,{\bf V}_i) &= 0,\\[0.5ex]
m_i \,n\,\frac{\partial  {\bf V}_i}{\partial t} +
m_i \,n\,({\bf V}_i\cdot\nabla) {\bf V}_i +
 [\delta_i^{-2}]\,\nabla p_i + 
[\zeta_i^{-1}]\,\nabla\cdot \bpi_i &\nonumber\\[0.5ex]
- [\delta_i^{-2}]\,e \,n\,
({\bf E} + {\bf V}_i\times {\bf B})& =-[\delta_i^{-2}\,\zeta_i\,\mu]\, {\bf F}_U - [\delta_i^{-2}]\,{\bf F}_T,
\label{e3.152b}\\[0.5ex]
\frac{3}{2}\frac{\partial  p_i}{\partial t} +   \frac{3}{2}
 \,{\bf V}_i\cdot\nabla p_i
+ \frac{5}{2}\,p_i\,\nabla\cdot{\bf V}_i+[\delta_i^{\,2}\,\zeta_i^{-1}]\, \bpi_i:\nabla{\bf V}_i\nonumber\\[0.5ex]
+ [\zeta_i^{-1}]\,\nabla\cdot{\bf q}_i &= [\delta_i^{-2}\,\zeta_i\,\mu]\,w_i.\label{e3.152c}
\end{align}
The terms in square brackets are there to remind us that the terms they precede are smaller or larger than the
other terms in the equations (by the corresponding terms within the brackets).

If we assume that $\delta_e\sim \zeta_e\sim \mu\ll 1$ then the dominant terms in the electron energy conservation equation
(\ref{e3.151c}) yield
\begin{equation}
\nabla\cdot{\bf q}_{T_e}\simeq \nabla\cdot(-\kappa_\parallel^e\,\nabla_\parallel T_e)\simeq 0,
\end{equation}
which implies that 
\begin{equation}
{\bf B}\cdot\nabla T_e=0. \label{e75}
\end{equation}
In other words, the parallel electron heat conductivity in a tokamak plasma is usually
sufficiently large that it forces the electron temperature to be constant on magnetic flux-surfaces. 
Likewise, if we assume that $\delta_i\sim \zeta_i\sim \mu\ll 1$ then the dominant terms in the ion energy conservation equation
(\ref{e3.152c}) yield
\begin{equation}
\nabla\cdot{\bf q}_{T_i}\simeq \nabla\cdot(-\kappa_\parallel^i\,\nabla_\parallel T_i)\simeq 0,
\end{equation}
which implies that 
\begin{equation}
{\bf B}\cdot\nabla T_i=0.\label{e77}
\end{equation}
In other words, the parallel ion heat conductivity in a tokamak plasma is usually
sufficiently large that it forces the ion temperature to be constant on magnetic flux-surfaces.

 The dominant terms in the electron and ion momentum conservation equations,
(\ref{e3.151b}) and (\ref{e3.152b}), respectively, yield
\begin{align}\label{e78}
\nabla p_e + n\,e\,({\bf E}+{\bf V}_e\times {\bf B}) &\simeq 0,\\[0.5ex]
\nabla p_i - n\,e\,({\bf E}+{\bf V}_i\times {\bf B}) &\simeq 0.\label{e79}
\end{align}
The sum of the preceding two equations gives
\begin{equation}\label{e80}
 {\bf j}\times {\bf B} \simeq\nabla p,
\end{equation}
where $p=p_e+p_i$. In other words, a conventional tokamak plasma exists in a state in which electromagnetic
forces exactly balance the  total scalar pressure force. 
It follows from the previous equation that 
\begin{equation}
{\bf B}\cdot\nabla p\simeq 0.
\end{equation}
In other words, the pressure in a conventional tokamak plasma is constant on magnetic flux-surfaces. 
 However, given that $p=n\,(T_e+T_i)$, and making use of Eqs.~(\ref{e75}) and (\ref{e77}), we deduce that 
\begin{equation}
{\bf B}\cdot\nabla n\simeq 0.
\end{equation}
In other words, the electron number density in a conventional tokamak plasma is also  constant on
magnetic flux-surfaces.

 Taking the difference of the electron and ion particle
conservation equations, (\ref{e3.151a}) and (\ref{e3.152a}), respectively, we obtain
\begin{equation}\label{e82}
\nabla\cdot {\bf j} = 0.
\end{equation}
 Equations~(\ref{e80}) and (\ref{e82}) can be combined to give
\begin{equation}\label{e84}
({\bf B}\cdot\nabla)\,{\bf j } - ({\bf j}\cdot\nabla)\,{\bf B} \simeq {\bf 0}.
\end{equation}
Here, we have made use of the fact that $\nabla\cdot{\bf B}=0$.

Finally, Equations~(\ref{e78}) and (\ref{e79}) imply that
\begin{align}
{\bf V}_{\perp\,e} &\simeq {\bf V}_E + {\bf V}_{\ast\,e},\\[0.5ex]
{\bf V}_{\perp\,i} &\simeq {\bf V}_E + {\bf V}_{\ast\,i},
\end{align}
where
\begin{equation}
{\bf V}_E = \frac{{\bf E}\times {\bf B}}{B^{\,2}}
\end{equation}
is the {\em E-cross-B drift velocity}\/ (otherwise known as  the ${\bf E}\times {\bf B}$ velocity), and
\begin{align}
{\bf V}_{\ast\,e}&= \frac{\nabla p_e\times {\bf B}}{e\,n\,B^{\,2}},\\[0.5ex]
{\bf V}_{\ast\,i}&=- \frac{\nabla p_i\times {\bf B}}{e\,n\,B^{\,2}},
\end{align}
are termed the {\em electron diamagnetic velocity}\/ and the {\em ion diamagnetic velocity}, respectively (Fitzpatrick 2015).

\section{Cylindrical Tearing Mode Theory}
\subsection{Introduction}
The aim of this section is to describe the simplest theory of tearing mode dynamics in tokamak plasmas, according to which the plasma
equilibrium is approximated as a periodic cylinder.

\subsection{Tokamak Equilibrium}
Consider a large aspect-ratio, toroidal,  tokamak plasma equilibrium whose magnetic flux-surfaces map out (almost) concentric circles in the poloidal plane. Such an equilibrium can be approximated as a periodic cylinder (Wesson 1978). 
Let us employ a conventional set of right-handed cylindrical coordinates, $r$, $\theta$, $z$. The
equilibrium magnetic flux-surfaces lie on surfaces of constant $r$. The system is
assumed to be periodic in the $z$ (`toroidal') direction, with periodicity length $2\pi\,R_0$, 
where $R_0$ is the simulated {\em major radius}\/ of the plasma. The {\em safety-factor}\/ profile takes the form
\begin{equation}\label{e90}
q(r)= \frac{r\,B_z}{R_0\,B_\theta(r)},
\end{equation}
where $B_z$ is the constant `toroidal' magnetic field-strength, and $B_\theta(r)$ is
the poloidal magnetic field-strength. It is assumed that $q\sim{\cal O}(1)$. 
The equilibrium `toroidal' current density
satisfies
\begin{equation}
\mu_0\,j_z(r) = \frac{(r\,B_\theta)'}{r},
\end{equation}
where $'$ denotes $d/dr$. Finally, the standard large aspect-ratio
tokamak orderings,  
\begin{equation}\label{e92}
\frac{r}{R_0}\ll 1,~~~ \frac{B_\theta}{B_z}\ll 1,
\end{equation}
are adopted (Fitzpatrick 1993).

\subsection{Perturbed Magnetic Field}
Consider a tearing mode perturbation that has $m$ periods in the poloidal direction,
and $n$ periods in the toroidal direction, where $m/n\sim {\cal O}(1)$.  We can write the perturbed magnetic field in the form
\begin{equation}\label{e93}
\delta{\bf B} \simeq \nabla\times (\psi\,{\bf e}_z),
\end{equation}
where
\begin{equation}\label{e93a}
\psi(r,\theta,\varphi,t)= \psi(r,t)\,\exp[{\rm i}\,(m\,\theta-n\,\varphi)].
\end{equation}
Here, $\varphi=z/R_0$ is a simulated toroidal angle. The perturbed current density becomes
\begin{equation}\label{e94}
\mu_0\,\delta {\bf j} \simeq - \nabla^2\psi\,{\bf e}_z.
\end{equation}
It is assumed that $|\delta{\bf B}|\ll |{\bf B}|$.

\subsection{Cylindrical Tearing Mode Equation}
To lowest order, the tearing perturbation is governed by the linearized version of Equation~(\ref{e84}), which yields
\begin{equation}\label{e96}
(\delta{\bf B}\cdot \nabla)\,{\bf j} + ({\bf B}\cdot\nabla)\,\delta{\bf j} - (\delta{\bf j}\cdot\nabla)\,{\bf B} - 
({\bf j}\cdot\nabla)\,\delta {\bf B}\simeq {\bf 0}.
\end{equation}
Making use of Equations~(\ref{e90})--(\ref{e94}), the $z$-component of the previous equation gives
the {\em cylindrical tearing mode equation}\/ (Fitzpatrick 1993),
\begin{equation}
\frac{\partial^2\psi}{\partial r^2} + \frac{1}{r}\,\frac{\partial\psi}{\partial r}-\frac{m^2}{r^2}\,\psi - \frac{J_z'\,\psi}{r\,(1/q-n/m)}\simeq 0,
\end{equation}
where 
\begin{equation}
J_z(r)= \frac{R_0\,\mu_0\,j_z(r)}{B_z}.
\end{equation}

\subsection{Solution in Presence of Perfectly-Conducting Wall}\label{perfect}
Suppose that the plasma occupies the region $0\leq r\leq a$, where $a$ is the plasma {\em minor radius}. It follows that
$J_z(r)=0$ for $r>a$. Let the plasma be surrounded by a concentric perfectly-conducting wall of radius $r_w>a$. 
(By definition, $a\ll R_0$ in a large aspect-ratio tokamak.) An appropriate
physical solution of the cylindrical tearing mode equation takes the form
\begin{equation}\label{e99}
\psi(r,t)= {\mit\Psi}_s(t)\,\hat{\psi}_s(r),
\end{equation}
where the real function $\hat{\psi}_s(r)$ is a solution of 
\begin{equation}\label{e100}
\frac{d^2\hat{\psi}_s}{dr^2} + \frac{1}{r}\,\frac{d\hat{\psi}_s}{dr}-\frac{m^2}{r^2}\,\hat{\psi}_s - \frac{J_z'\,\hat{\psi}_s}{r\,(1/q-n/m)}\simeq 0
\end{equation}
that satisfies
\begin{align}\label{e101}
\hat{\psi}_s(0) &= 0,\\[0.5ex]
\hat{\psi}_s(r_s) &= 1,\label{e102}\\[0.5ex]
\hat{\psi}_s(r\geq r_w) &= 0.\label{e103}
\end{align}
